---
title: "Analysing time-course (phospho)proteomic data"
author: "Pengyi Yang"
date: "Dec 2019"
output:
  html_document:
    code_folding: show
    fig_height: 8
    fig_width: 8
    toc: yes
    number_sections: true
    toc_depth: 3
    toc_float: yes
---

# Introduction

While 1, 2, and 3D pathway analyses are useful for data generated from experiments with different treatment/conditions, analysis designed for time-course data may be better suited to analysis experiments that profile multiple time points such as in the ESC differentiation experiment.

Here, we will introduce `ClueR` which is an R package specifically designed for time-course proteomic and phosphoproteomic data analysis ([Pengyi Yang et al. 2015])(https://journals.plos.org/ploscompbiol/article?id=10.1371/journal.pcbi.1004403).

Let us start with gene-centric analysis. Such analysis can be directly applied to proteomics data. It can also be applied to phosphoproteomic data by using the `phosCollapse` function to summarise phosphosite information to the gene level.

```{r}
suppressPackageStartupMessages({
  library(PhosR)
  library(ClueR)
  library(ggplot2)
  library(parallel)
})
```

```{r}
# load example data
data("phosphoESC")

# summarise phosphosite information to the gene level
phosphoESC.sum <- phosCollapse(phosphoESC, gsub(";.+", "", rownames(phosphoESC)), stat = apply(abs(phosphoESC), 1, max), by = "max")
rownames(phosphoESC.sum) <- toupper(rownames(phosphoESC.sum))
```

```{r, eval = FALSE}
# NOT RUN. to save the time the following block is not run.
# perform ClueR to identify optimal number of clusters.
RNGkind("L'Ecuyer-CMRG")
set.seed(123)
c1 <- runClue(phosphoESC.sum, annotation=Pathways.reactome, kRange = 2:30, rep = 5, effectiveSize = c(5, 100), pvalueCutoff = 0.05, alpha = 0.5)
```

Visualise the evaluation results

```{r}
load("phosphoESC_cluster.RData")

data <- data.frame(Success=as.numeric(c1$evlMat), Freq=rep(2:30, each=5))
myplot <- ggplot(data, aes(x=Freq, y=Success)) + geom_boxplot(aes(x = factor(Freq), fill="gray"))+
  stat_smooth(method="loess", colour="red", size=3, span = 0.5) + xlab("# of cluster")+ ylab("Enrichment score")+theme_classic()
myplot
```

Finding enriched pathways from each cluster.

```{r, fig.height=10, fig.width=10}
set.seed(123)
best <- clustOptimal(c1, rep=5, mfrow=c(4, 4), visualize = TRUE)
best$enrichList
```



# Session Info
```{r}
sessionInfo()
```




